---
title: "Pulling and Processing Historical Raw Population Data"
author: "Jaspreet Kang"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Introduction

Note: This Rmarkdown only needs to be executed once to produce the historical population datasets. Once produced, this file never needs to be ran again **unless** any of the historical data files (listed further below) are updated after April 3, 2024.

This RMarkdown file extracts and processes historical population data necessary for creating sex-age-race/ethnicity estimates for all 61 California Local Health Jurisdictions:

-   58 counties
-   3 city-based jurisdictions: Berkeley, Long Beach, and Pasadena
-   Alameda LHJ
-   Los Angeles LHJ

## Population Data Required:

The population data files extracted and processed include:

1.  Department of Finance (DOF) Annual Intercensal Population Estimates by Race/Ethnicity with Age and Gender Detail

    1.  Downloaded from [here](https://dof.ca.gov/forecasting/demographics/estimates/annual-intercensal-population-estimates-by-race-ethnicity-with-age-and-gender-detail/)

    2.  Location: *P3/Intercensal_2000-2010_DBInput.csv*

    3.  Contains county-level Sex-Age-Race/Ethnicity population estimates for April 1, 2000 and 2010 and July 1, 2000 through 2010 estimates. July 1, 2000 through July 1, 2009 estimates are used.

2.  Census Decennial for county- and city-level Sex-Age-Race/Ethnicity population (April 1st estimates)

    1.  Extracted using R's [tidycensus](https://github.com/walkerke/tidycensus) package. This method requires a Census API key, which can be obtained [here](https://api.census.gov/data/key_signup.html). Estimates are pulled from the 2000 and 2010 Decennial.

### Optional

Department of Finance P3 data (processed in *1b-Process-Current-Raw-Data.Rmd*) are January 1 estimates. The files below allow for the adjustment of DOF P3 data to July 1 population rather than January 1.

3.  Department of Finance - E-4 Historical Population Estimates for Cities, Counties, and the State

    1.  Excel files of these data were downloaded from [here](https://dof.ca.gov/forecasting/demographics/estimates/), and are stored in *E4/Raw/*. The raw data were then manually formatted for easier processing and stored into *E4/Clean/*.

    2.  Contains January 1st county- and city-level total estimates from 1970-2010

4.  Department of Finance - E-6 Population Estimates and Components of Change by County

    1.  Excel files of these data were downloaded from [here](https://dof.ca.gov/forecasting/demographics/estimates/), and are stored in *E6/Raw/*. The raw data were then manually formatted for easier processing and stored into *E6/Clean/.*

    2.  Contains July 1st county-level total estimates from 1947-2009

Note: The data structure and formatting between each of DOF's raw E-4 and E-6 files are largely inconsistent. Therefore, these files were first manually converted to a standard format.

## Data File Outputs:

The processed datasets are saved as csv and RDS files into *intermediatePopData/*:

1.  *dof-p3-processed-historical*

    1.  Note: This file is processed data of the DOF annual intercensal sex-age-race/ethnicity population estimates (from 2000-2009) at the county-level. We call it "P3" for convenience, since this data is eventually combined with the current DOF P3 data which also contains sex-age-race/ethnicity population estimates at the county-level. One major difference is the 2000-2009 annual intercensal data are July 1st estimates, whereas the P3 data are Jan 1st estimates.

2.  *census-city-ars-historical*

3.  *census-county-ars-historical*

4.  *dof-e4-processed-historical*

5.  *dof-e6-processed-historical*

# Processing

## Setup

R version 4.0.4 (2021-02-15)

Platform: x86_64-pc-linz-gnu (64-bit)

RStudio Workbench Version 1.4.1717-3: "Juliet Rose" (23fd1677, 2021-05-27) for CentOS 8

Package versions:

-   readxl - 1.3.1

-   readr - 2.1.4

-   dplyr - 1.1.3

-   tidyr - 1.3.0

-   tidycensus - 1.5

-   janitor - 2.1.0

-   stringr - 1.5.0

-   lubridate - 1.9.3

```{r}

# Setup ========================================================================================

# Disable scientific notation
options(scipen=999)

## Load packages ----------------------------------------------
library(readxl) # Read in excel files
library(readr) # Read in and write csv files
library(dplyr) # Data manipulation
library(tidyr) # Data reshaping
library(janitor) # Contains helpful functions for data cleaning
library(lubridate) # Date formatting
library(stringr) # String formatting
library(tidycensus) # For interacting with Census API

## Global constants ---------------------------------------------

# City LHJs
cities <- c("Berkeley", "Long Beach", "Pasadena")


# Directory to save output files to
outputDir <- "intermediatePopData/"

```

## Process and Save Data

### (Required) Department of Finance's **Annual Intercensal Population Estimates by Race/Ethnicity with Age and Gender Detail**

Steps:

1.  Read in relationship files, located in *Standards/*

    1.  *countyLink.xlsx* - Used for linking DOF's county codes to county names.

    2.  *raceLink.xlsx* - Used for linking DOF's race/ethnicity codes to a standard set of race/ethnicity labels.

2.  Read in DOF's "Annual Intercensal Population Estimates by Race/Ethnicity with Age and Gender Detail" file

    1.  Location: *P3/Intercensal_2000-2010_DBInput.csv*

    2.  Contains Sex-Age-Race/Ethnicity population estimates for April 1, 2000 and 2010 and July 1, 2000 through 2010 estimates. July 1, 2000 through July 1, 2009 estimates are used.

3.  Process data

4.  Save data as .RDS and .csv into *intermediatePopData/*

    1.  *dof-p3-processed-historical.RDS*

    2.  *dof-p3-processed-historical.csv*

```{r}

# County relationship file =======
countyLink <- read_xlsx("Standards/countyLink.xlsx") %>% 
  select(county = countyName, CountyCode = cdphcaCountyTxt)

raceLink <- read_xlsx("Standards/raceLink.xlsx") %>% 
  select(race7, raceEth = raceNameShort)

# Read in and process data =======
p3_historical <- read_csv("P3/Intercensal_2000-2010_DBInput.csv") %>% 
  mutate(year  = as.numeric(str_sub(Year,5,9)),
         month = as.numeric(str_sub(Year,1,1))) %>%    
  filter(month == 7)  %>% # Use July estimates
  select(CountyCode, year, sex=Gender, race7=RaceCode, age=Age, population=Population) %>% # CountyCode - 2 digit character: # "01", "02" ... "58", "59"
  full_join(countyLink,by="CountyCode") %>%                       
  filter(CountyCode != "59") %>% # Remove California
  select(-CountyCode) %>%
  mutate(race7 = ifelse(race7 == 6, 99, race7), 
         race7 = ifelse(race7 == 7, 6, race7),
         race7 = ifelse(race7 == 99, 7, race7),
         age = as.character(age), 
         age = ifelse(age == "100", "100+", age)
         )  %>%
  left_join(raceLink, by = "race7") %>% 
  select(county, year, sex, raceEth, age, population) %>% 
  filter(year != 2010) # Remove 2010 here?

# Save data =======
saveRDS(p3_historical, paste0(outputDir, "dof-p3-processed-historical.RDS"))
write_csv(p3_historical, paste0(outputDir, "dof-p3-processed-historical.csv"))

```

### (Required) Census Decennial for county and city-level Sex-Age-Race/Ethnicity population

Steps:

1.  Load in Census API key into session. An API key can be obtained from [here](https://api.census.gov/data/key_signup.html).

    1.  Note: Set the **.ckey** object to your Census API key.

2.  Read in decennial relationship files, which link table IDs to their corresponding sex by age by R/E stratas

3.  Call the Census API to pull county and city-level sex by age by race/ethnicity estimates from the 2000 and 2010 Decennial

4.  Process outputs from Step 3

5.  Save data as .RDS and .csv into *intermediatePopData/*

    1.  *census-city-ars-historical.RDS*

    2.  *census-city-ars-historical.csv*

    3.  *census-county-ars-historical.RDS*

    4.  *census-county-ars-historical.csv*

```{r}

# Census API Key: Set your API key here, which can be obtained from
# https://api.census.gov/data/key_signup.html
.ckey   <- read_file("Standards/census.api.key.txt")
census_api_key(.ckey) # Load census API key into sesesion

# Read in Census Decennial relationship file ======================
# Links table IDs to their corresponding sex by age by R/E stratas

decLink2000 <- read_excel("Standards/decennialLink.xlsx", sheet = "2000 ARS")
decLink2010 <- read_excel("Standards/decennialLink.xlsx", sheet = "2010 ARS")

# Custom Function to pull decennial census data =============================
# Calls tidycensus' get_decennial() function, and links the resultant data frame to our census dcennial relationship file

# Function arguments:
# - myYear (num): a decennial year
# - myGeography (chr): "place" for city; "county" for county
# - mySumFil (chr)e: "sf1" for Decennial 2000 and 2010
# - cache_table (TRUE or FALSE): Whether or not to cache table names for faster future access. Defaults to FALSE; if TRUE, only needs to be called once per dataset
pullDecennial <- function(myYear, myGeography, mySumFile = "sf1", cacheTable = F) {
  
  if (myYear == 2000) {
    decLink <- decLink2000
  } else if (myYear == 2010) {
    decLink <- decLink2010
  } 
  
  tableIDs <- unique(decLink$tableID)
  
  lapply(tableIDs, function(x, myYear1 = myYear) {
    
    tDat <- get_decennial(geography = myGeography, table = x, year = myYear1, state = 06, sumfile = mySumFile, cache_table = cacheTable) %>% 
      left_join(select(decLink, -tableID), by = c("variable" = "name")) %>% 
      mutate(year = myYear1) %>% 
      select(year, GEOID, NAME, sex, age, raceEth, population = value)
  
  
}) %>% 
  bind_rows()
}


## Call function pullDecennial() to pull 2000 and 2010 Decennial Census data  ========
census2000_ars_city <- pullDecennial(myYear = 2000, myGeography = "place")
census2000_ars_county <- pullDecennial(myYear = 2000, myGeography = "county")

census2010_ars_city <- pullDecennial(myYear = 2010, myGeography = "place")
census2010_ars_county <- pullDecennial(myYear = 2010, myGeography = "county")

## Bind, process, and save data =======================
censusFinal_city <- bind_rows(census2000_ars_city, census2010_ars_city) %>% 
  group_by(year, GEOID, NAME, sex, age, raceEth) %>% 
  summarise(population = sum(population)) %>% 
  ungroup()

censusFinal_county <- bind_rows(census2000_ars_county, census2010_ars_county) %>% 
  group_by(year, GEOID, NAME, sex, age, raceEth) %>% 
  summarise(population = sum(population)) %>% 
  ungroup()

saveRDS(censusFinal_city, paste0(outputDir, "census-city-ars-historical.RDS"))
write_csv(censusFinal_city, paste0(outputDir, "census-city-ars-historical.csv"))

saveRDS(censusFinal_county, paste0(outputDir, "census-county-ars-historical.RDS"))
write_csv(censusFinal_county, paste0(outputDir, "census-county-ars-historical.csv"))
```

### (Optional) Department of Finance - E-4 Historical Population Estimates for Cities, Counties, and the State

Steps:

1.  Read in files from *E4/Clean/*

    1.  *E4-1971-1980.xlsx*

    2.  *E4-1981-1990.xlsx*

    3.  *E4-1991-2000.xlsx*

    4.  *E4-2001-2010.xlsx*

2.  Data processing which includes

    1.  Filtering the city portion of the data on Berkeley, Long Beach, and Pasadena

    2.  Filtering the county portion of the data on Alameda County and Los Angeles County

    3.  Calculating city to county proportion

        1.  Berkeley population / Alameda County population

        2.  Pasadena population / Los Angeles County population

        3.  Long Beach population / Los Angeles County population

3.  Save data as .RDS and .csv into *intermediatePopData/*

    1.  *dof-e4-processed-historical.RDS*

    2.  *dof-e4-processed-historical.csv*

```{r}

# Function for processing E4 files -----------------
process_e4 <- function(years) { 
  
  colNames <- c("place", paste0("1/1/", years))
  
  fileName <- ifelse(2021 %in% years, 
                     "E4/Clean/E4-2021-cy.xlsx",
                     paste0("E4/Clean/E4-", min(years), "-", max(years), ".xlsx"))
  
  tCountyCity <- lapply(c("County", "City"), function(x) {
    
    read_excel(fileName, skip = 1, col_names = colNames, sheet = x) %>% 
      select(place, starts_with("1")) %>% 
      mutate(across(starts_with("1"), ~as.numeric(.x))) %>% 
      pivot_longer(-place, names_to = "year", values_to = "population") %>% 
      mutate(year = as.numeric(substr(year, 5, 8)))
    
  })
  names(tCountyCity) <- c("County", "City")
  
  tCity <- tCountyCity$City %>% 
    filter(place %in% cities) %>% 
    mutate(county = ifelse(place == "Berkeley", "Alameda County", "Los Angeles County")) %>% 
    rename(city_population = population) %>% 
    left_join(rename(tCountyCity$County, county = place, county_population = population)) %>% 
    mutate(city_county_proportion = city_population / county_population, 
           county = sub(" County.*", "", county)) %>% 
    select(year, place, county, city_county_proportion)
  
}

# Process and save ---------------------------
e4_final <- bind_rows(
  process_e4(1971:1980), 
  process_e4(1981:1990),
  process_e4(1991:2000),
  process_e4(2001:2010)
)

saveRDS(e4_final, paste0(outputDir, "dof-e4-processed-historical.RDS"))
write_csv(e4_final, paste0(outputDir, "dof-e4-processed-historical.csv"))

```

### (Optional) Department of Finance - E-6 Population Estimates and Components of Change by County

Steps:

1.  Read in files from *E6/Clean/*

    1.  *E6-1947-1969.xlsx*

    2.  *E6-1970-1989.xlsx*

    3.  *E6-1990-1999.xlsx*

    4.  *E6-2000-2009.xlsx*

2.  Process data

3.  Save data as .RDS and .csv into *intermediatePopData/*

    1.  *dof-e6-processed-historical.RDS*

    2.  *dof-e6-processed-historical.csv*

```{r}

# Process 1947-1969 data --------------------------------

e6_1947_69 <- read_excel("E6/Clean/E6-1947-1969.xlsx", col_names = c("county", 1947:1969), skip = 1) %>% 
  pivot_longer(-county, names_to = "year", values_to = "population") %>% 
  mutate(year = as.numeric(year))

# Function for processing 1970-2009 data --------------------
process_e6 <- function(years) {
  
  fileName <- ifelse(2021 %in% years, 
                     "E6/Clean/E6-2020-cy.xlsx",
                     paste0("E6/Clean/E6-", min(years), "-", max(years), ".xlsx"))
  
  read_excel(fileName) %>% 
    rename(county = County, year = Year) %>% 
    mutate(year = as.numeric(year), 
           population = as.numeric(population))
  
}

# Process 1970-CY data -----------------------------------
e6_final <- bind_rows(e6_1947_69, 
                      process_e6(1970:1989), 
                      process_e6(1990:1999),
                      process_e6(2000:2009)
                      )

# Save data --------------------------------------
saveRDS(e6_final, paste0(outputDir, "dof-e6-processed-historical.RDS"))
write_csv(e6_final, paste0(outputDir, "dof-e6-processed-historical.csv"))

```
