---
title: "Pulling and Processing Recent Raw Population Data"
author: "CDPH/LHJ Population Task Force"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, eval = FALSE)
```

# Introduction

This RMarkdown file extracts and processes current population data.

Note: This Rmarkdown is largely the same as *1a-Process-Historical-Raw-Data.Rmd*, except that it processes recent and current population datasets. Therefore, this file must be maintained, updated, and executed whenever a new population data file (either from Department of Finance or Census Decennial) is released.

## Population Data Required:

The population data files extracted and processed include:

1.  Department of Finance (DOF) Annual Intercensal Population Estimates by Race/Ethnicity with Age and Gender Detail

    1.  Location: *P3/intercensals_2010_2020.csv*

    2.  Released 2025.01.29

    3.  Downloaded from [here](https://dof.ca.gov/forecasting/demographics/estimates/annual-intercensal-population-estimates-by-race-ethnicity-with-age-and-gender-detail-2010-2020/)

    4.  2010-2019 population estimates are used.

    5.  Contains sex by age by race/ethnicity population estimates at the county-level.

2.  Department of Finance (DOF) P3 Complete State and County Projections Dataset

    1.  Location: *P3/P3_Complete.csv*

    2.  Vintage 2025 (released 2025.04.25)

    3.  Downloaded from [here](https://dof.ca.gov/forecasting/demographics/projections/)

    4.  2020-2024 population estimates are used.

    5.  Contains sex by age by race/ethnicity population estimates at the county-level.

3.  2020 Census Decennial for county- and city-level Sex-Age-Race/Ethnicity population

    1.  Extracted using R's [tidycensus](https://github.com/walkerke/tidycensus) package. This method requires a Census API key, which can be obtained [here](https://api.census.gov/data/key_signup.html). Estimates are pulled from the 2020 Decennial.

4.  Department of Finance - E-4 Population Estimates for Cities, Counties, and the State

    1.  Excel files of these data were downloaded from [here](https://dof.ca.gov/forecasting/demographics/estimates/), and are stored in *E4/Raw/*. The raw data were then manually formatted for easier processing and stored into *E4/Clean/*.

    2.  Contains January 1st county- and city-level total estimates from 2011-2025

5.  Department of Finance - E-6 Population Estimates and Components of Change by County

    1.  Excel files of these data were downloaded from [here](https://dof.ca.gov/forecasting/demographics/estimates/), and are stored in *E6/Raw/*. The raw data were then manually formatted for easier processing stored into *E6/Clean/.*

    2.  Contains July 1st county-level total estimates from 2010-2024

Note: The data structure and formatting between each of DOF's raw E-4 and E-6 files are largely inconsistent. Therefore, these files were first manually converted to a standard format.

## Data File Outputs:

The processed datasets are saved as csv and RDS files into *intermediatePopData/*:

1.  *dof-p3-processed-current*
2.  *census-city-ars-current*
3.  *census-county-ars-current*
4.  *dof-e4-processed-current*
5.  *dof-e6-processed-current*

# Processing

## Setup

Note: The final product (LHJ-level sex-age-race/ethnicity population estimates - created in *3-Create-LHJ-Population-ARS.Rmd*) spans from 2000 to the most recent year of data available in DOF-E6 (**currentYear** object below).

```{r}

# Setup ========================================================================================

# Disable scientific notation
options(scipen=999)

## Load packages ----------------------------------------------
library(readxl) # Read in excel files
library(readr) # Read in and write csv files
library(dplyr) # Data manipulation
library(tidyr) # Data reshaping
library(janitor) # Contains helpful functions for data cleaning
library(lubridate) # Date formatting
library(stringr) # String formatting
library(tidycensus) # For interacting with Census API

## Global constants ---------------------------------------------

# Set current year
currentYear <- 2024 # Jan E6 - most recent year available
currentYear_Jan <- 2025 # Jan E4 - most recent year available

# City LHJs
cities <- c("Berkeley", "Long Beach", "Pasadena")


# Directory to save output files to
outputDir <- "intermediatePopData/"

```

## Process and Save Data

### Department of Finance's P3 Complete State and County Projections Dataset

```{r}

# County relationship file: Used for linking DOF's county codes to county names. =======
countyLink <- read_xlsx("Standards/countyLink.xlsx") %>% 
  select(county = countyName, fips = FIPSCounty) %>% 
  mutate(fips = as.numeric(paste0("6", fips)))

# Race labels crosswalk: Used for linking DOF's race/ethnicity codes to a standard set of race/ethnicity labels.
raceLink <- read_xlsx("Standards/raceLink.xlsx") %>% 
  select(race7, raceEth = raceNameShort) %>% 
  filter(!is.na(race7))

# Read in and process data =======

if (F) {
  # Previous DOF P3 ---------------------
  # Vintage 2020 (2021.7.14) version
  p3_10_60 <- read_csv("P3/P3_Complete_2010-2060.csv") %>%
    left_join(countyLink, by = "fips") %>% 
    mutate(sex = str_to_title(sex)) %>% 
    select(county, year, sex, race7, age = agerc, population = perwt)
  
  # 2010-currentYear
  p3_10_ <- p3_10_60 %>% 
    filter(year %in% 2010:currentYear) %>% 
    mutate(source = "old dof")
  
  
  # 2020 - Current --------------
  # Vintage 2023 (2024.3.8) version
  p3_20_ <- read_csv("P3/P3_Complete_2020-2060.csv") %>% 
    filter(year %in% 2020:currentYear) %>% 
    left_join(countyLink, by = "fips") %>% 
    mutate(sex = str_to_title(sex)) %>% 
    select(county, year, sex, race7, age = agerc, population = perwt) %>% 
    mutate(source = "new dof")

  # Bind data ===================
  p3_final <- bind_rows(p3_10_, p3_20_) %>% 
    left_join(raceLink) %>% 
    select(-race7) %>% 
    mutate(age = case_when(age >= 100 ~ "100+", 
                           TRUE ~ as.character(age))) %>% 
    group_by(county, year, sex, age, raceEth, source) %>% 
    summarise(population = sum(population)) %>% 
    ungroup()
  
}

# DOF Intercensal 2010-2019
intercensal_10_19 <- read_csv("P3/intercensals_2010_2020.csv") %>% 
  filter(month == 7) %>% 
  left_join(countyLink, by = "fips") %>% 
  left_join(raceLink) %>% 
  mutate(sex = str_to_title(sex)) %>% 
  select(county, year, sex, raceEth, age = agerc, population = perwt)

# DOF P3 Vintage 2025 (2025.04.25 version)
p3_20_ <- read_csv("P3/P3_Complete.csv") %>% 
  filter(year %in% 2020:currentYear) %>% 
  left_join(countyLink, by = "fips") %>% 
  left_join(raceLink) %>%
  mutate(sex = str_to_title(sex)) %>% 
  select(county, year, sex, raceEth, age = agerc, population = perwt)

# Bind data ===================
p3_final <- bind_rows(intercensal_10_19, p3_20_) %>% 
  mutate(age = case_when(age >= 100 ~ "100+", 
                         TRUE ~ as.character(age))) %>% 
  group_by(county, year, sex, age, raceEth) %>% 
  summarise(population = sum(population)) %>% 
  ungroup()


# Save data =======
saveRDS(p3_final, paste0(outputDir, "dof-p3-processed-current.RDS"))
write_csv(p3_final, paste0(outputDir, "dof-p3-processed-current.csv"))

```

### Census Decennial for county and city-level Sex-Age-Race/Ethnicity population

Note: Census API key required, which can be obtained from [here](https://api.census.gov/data/key_signup.html).

```{r}

# Census API Key: Set your API key here, which can be obtained from
# https://api.census.gov/data/key_signup.html
.ckey   <- read_file("Standards/census.api.key.txt")
census_api_key(.ckey) # Load census API key into sesesion

# Read in Census Decennial relationship file ======================
# Links table IDs to their corresponding sex by age by R/E stratas

decLink2020 <- read_excel("Standards/decennialLink.xlsx", sheet = "2020 ARS")

# Custom Function to pull decennial census data =============================
# Calls tidycensus' get_decennial() function, and links the resultant data frame to our census dcennial relationship file

# Function arguments:
# - myYear (num): 2020 is only option
# - myGeography: "place" for city; "county" for county
# - mySumFile: "dhc" for Decennial 2020
# - cache_table (TRUE or FALSE): Whether or not to cache table names for faster future access. Defaults to FALSE; if TRUE, only needs to be called once per dataset

pullDecennial <- function(myYear = 2020, myGeography, mySumFile = "dhc", cacheTable = FALSE) {
  
  if (myYear == 2020) {
    decLink <- decLink2020
  } else {
    stop("Error: Pass in valid year")
  } 
  
  tableIDs <- unique(decLink$tableID)
  
  lapply(tableIDs, function(x, myYear1 = myYear) {
    
    tDat <- get_decennial(geography = myGeography, table = x, year = myYear1, state = 06, sumfile = mySumFile, cache_table = cacheTable) %>% 
      left_join(select(decLink, -tableID), by = c("variable" = "name")) %>% 
      mutate(year = myYear1) %>% 
      select(year, GEOID, NAME, sex, age, raceEth, population = value)
  
  
}) %>% 
  bind_rows()
}

# Get total first
decLink2020_total <- decLink2020 %>% 
  filter(sex == "Total", age == "Total", raceEth == "Total") %>% 
  pull(name)

census2020_total_city <- get_decennial(geography = "place", year = 2020, variables = decLink2020_total, state = 06) %>% 
  mutate(sex = "Total", age = "Total", raceEth = "Total", year = 2020) %>% 
  select(year, GEOID, NAME, sex, age, raceEth, population = value)

census2020_total_county <- get_decennial(geography = "county", year = 2020, variables = decLink2020_total, state = 06) %>% 
  mutate(sex = "Total", age = "Total", raceEth = "Total", year = 2020) %>% 
  select(year, GEOID, NAME, sex, age, raceEth, population = value)


## Call function pullDecennial() to pull 2020 Decennial Census data  ========
decLink2020 <- decLink2020 %>% 
  filter(!is.na(tableID))

census2020_ars_city <- pullDecennial(myYear = 2020, myGeography = "place", mySumFile = "dhc")
census2020_ars_county <- pullDecennial(myYear = 2020, myGeography = "county", mySumFile = "dhc")

## Bind, process, and save data =======================
censusFinal_city <- census2020_ars_city %>% 
  group_by(year, GEOID, NAME, sex, age, raceEth) %>% 
  summarise(population = sum(population)) %>% 
  ungroup() %>% 
  bind_rows(census2020_total_city)

censusFinal_county <- census2020_ars_county %>% 
  group_by(year, GEOID, NAME, sex, age, raceEth) %>% 
  summarise(population = sum(population)) %>% 
  ungroup() %>% 
  bind_rows(census2020_total_county)

saveRDS(censusFinal_city, paste0(outputDir, "census-city-ars-current.RDS"))
write_csv(censusFinal_city, paste0(outputDir, "census-city-ars-current.csv"))

saveRDS(censusFinal_county, paste0(outputDir, "census-county-ars-current.RDS"))
write_csv(censusFinal_county, paste0(outputDir, "census-county-ars-current.csv"))
```

### Department of Finance - E-4 Historical Population Estimates for Cities, Counties, and the State

1.  Data processing steps include:

    1.  Filtering the city portion of the data on Berkeley, Long Beach, and Pasadena

    2.  Filtering the county portion of the data on Alameda County and Los Angeles County

    3.  Calculating city to county proportion

        1.  Berkeley population / Alameda County population

        2.  Pasadena population / Los Angeles County population

        3.  Long Beach population / Los Angeles County population

```{r}

# Function for processing E4 files -----------------
process_e4 <- function(years) { 
  
  # Test
  if (F) {
    years <- 2011:2020
  }
  
  colNames <- c("place", paste0("1/1/", years))
  
  fileName <- ifelse(2021 %in% years, 
                     "E4/Clean/E4-2021-cy.xlsx",
                     paste0("E4/Clean/E4-", min(years), "-", max(years), ".xlsx"))
  
  tCountyCity <- lapply(c("County", "City"), function(x) {
    
    if (F) {
      x <- "County"
    }
    
    read_excel(fileName, skip = 1, col_names = colNames, sheet = x)%>% 
      select(place, starts_with("1")) %>% 
      mutate(across(starts_with("1"), ~as.numeric(.x))) %>% 
      pivot_longer(-place, names_to = "year", values_to = "population") %>% 
      mutate(year = as.numeric(substr(year, 5, 8)))
    
  })
  names(tCountyCity) <- c("County", "City")
  
  tCity <- tCountyCity$City %>% 
    filter(place %in% cities) %>% 
    mutate(county = ifelse(place == "Berkeley", "Alameda", "Los Angeles")) %>% 
    rename(city_population = population) %>% 
    left_join(rename(tCountyCity$County, county = place, county_population = population)) %>% 
    mutate(city_county_proportion = city_population / county_population, 
           county = sub(" County.*", "", county)) %>% # In case 'County' label exists
    select(year, place, county, city_county_proportion)
  
}

# Process and save ---------------------------
e4_final <- bind_rows(
  process_e4(2011:2020),
  process_e4(2021:currentYear_Jan)
) %>% 
  filter(year <= currentYear) # Restrict to most recent year available in July E6

saveRDS(e4_final, paste0(outputDir, "dof-e4-processed-current.RDS"))
write_csv(e4_final, paste0(outputDir, "dof-e4-processed-current.csv"))

```

### Department of Finance - E-6 Population Estimates and Components of Change by County

```{r}

# Function for processing 1970-2009 data --------------------
process_e6 <- function(years) {
  
  fileName <- ifelse(2021 %in% years, 
                     "E6/Clean/E6-2020-cy.xlsx",
                     paste0("E6/Clean/E6-", min(years), "-", max(years), ".xlsx"))
  
  read_excel(fileName) %>% 
    rename(county = County, year = Year) %>% 
    mutate(year = as.numeric(year), 
           population = as.numeric(population))
  
}

# Process 1970-CY data -----------------------------------
e6_final <- bind_rows(process_e6(2010:2019),
                      process_e6(2020:currentYear)
                      )

if (F) {
  e6_recent <- e6_final %>% 
    filter(year == currentYear-1) %>% 
    mutate(year = currentYear)
  
  e6_final <- e6_final %>% 
    bind_rows(e6_recent)
}

# Save data --------------------------------------
saveRDS(e6_final, paste0(outputDir, "dof-e6-processed-current.RDS"))
write_csv(e6_final, paste0(outputDir, "dof-e6-processed-current.csv"))

```
