---
title: "Create Total LHJ Population Data"
author: "CDPH/LHJ Population Data Work Group (PopDaWG)"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, eval = FALSE)
```

# Introduction

This RMarkdown file creates a dataset with total July LHJ population estimates. This dataset is used in *3-Create-LHJ-Population-ARS.Rmd* to perform the step of adjusting the DOF age-race-sex estimates to these estimates.

## Population Data Required

The following outputs are used from *1a-Process-Historical-Raw-Data.Rmd* and *1b-Process-Current-Raw-Data.Rmd*:

1.  *outputs/intermediate/dof-e4-processed-historical.RDS*

    1.  Contains city to county proportions (using January 1st estimates) from 1970-2010.

2.  *outputs/intermediate/dof-e4-processed-current.RDS*

    1.  Contains city to county proportions (using January 1st estimates) from 2011+.

3.  *outputs/intermediate/dof-e6-processed-historical.RDS*

    1.  Processed DOF July 1st estimates for counties from 1947-2009

4.  *outputs/intermediate/dof-e6-processed-current.RDS*

    1.  Processed DOF July 1st estimates for counties from 2010+
    
5.  *outputs/intermediate/dof-p3-processed-current.RDS*

    1.  Processed DOF July 1st estimates/projections for counties from 2020+. Only projection years are used in this script.

## Method

To calculate total July population estimates and projections for the 3 cities, multiply each city-to-county proportion to its respective county July population to get a city July population estimate and projection.

Then, to calculate total July population estimates/projections for Alameda and Los Angeles Health Departments:

1.  Alameda HD July population = Alameda County July population - Berkeley City July population

2.  Los Angeles HD July population = Los Angeles County July population - (Long Beach City July population + Pasadena City July population)

## Data File Outputs

Output is saved as a csv and RDS file into *outputs/intermediate/*:

1.  *lhj-population-total.RDS*
2.  *lhj-population-total.csv*

# Setup

```{r}


# Disable scientific notation
options(scipen=999) 

## Load packages ----------------------------------------------
library(readxl) # Read in excel files
library(readr) # Read in and write csv files
library(dplyr) # Data manipulation
library(tidyr) # Data reshaping

# City LHJs
cities <- c("Berkeley", "Long Beach", "Pasadena")

# Directory to read input data files from
inputDir <- "outputs/intermediate/"

# Directory to save output files to
outputDir <- inputDir

```

# Read in data

Read in the processed DOF E-4 and E-6 data - both historical and current.

```{r}

e4_historical <- readRDS(paste0(inputDir, "dof-e4-processed-historical.RDS"))
e4_current <- readRDS(paste0(inputDir, "dof-e4-processed-current.RDS"))
e4 <- bind_rows(e4_historical, e4_current) # Combine e4 historical and current


e6_historical <- readRDS(paste0(inputDir, "dof-e6-processed-historical.RDS"))
e6_current <- readRDS(paste0(inputDir, "dof-e6-processed-current.RDS"))
e6 <- bind_rows(e6_historical, e6_current) # Combine e6 historical and current

# P3 current - Projection years only
currentYear <- max(e6$year)
projectionYears <- (currentYear+1):(currentYear+5)

p3_current <- readRDS(paste0(inputDir, "dof-p3-processed-current.RDS")) %>% 
  filter(year %in% projectionYears) %>% 
  bind_rows(mutate(., county = "California")) %>% 
  group_by(year, county) %>% 
  summarise(population = sum(population)) %>% 
  ungroup()

# Add projection years to e4 and e6
# For e4: Use currentYear city:county proportion for projection years
# For e6: Use p3's projections

## E4 -----
currentYear_e4 <- max(e4$year)
addYears_e4 <- setdiff(projectionYears, currentYear_e4)

for (projectionYear in projectionYears) {
  e4 <- bind_rows(
    e4, 
    e4 %>% filter(year == currentYear_e4) %>% mutate(year = projectionYear)
  )
}

## E6 -----

e6 <- e6 %>% 
  bind_rows(p3_current)

```

# Calculate LHJ Total Estimates

```{r}

# County Estimates ---------------------------------------

countyEstimates <- e6

# City Estimates -----------------------------------------

cityEstimates <- e4 %>% 
  left_join(rename(e6, county_population = population)) %>% 
  mutate(city_population = city_county_proportion * county_population)


# Health Department Estimates ----------------------------
  
hdEstimates <- cityEstimates %>%
  mutate(place = ifelse(place == "Berkeley", "Berkeley", "Pasadena/Long Beach")) %>% 
  group_by(place, year, county) %>% 
  summarise(county_population = mean(county_population), 
            city_population = sum(city_population)) %>% 
  ungroup() %>% 
  mutate(population = county_population - city_population, 
         county_lhj = ifelse(place == "Berkeley", "Alameda HD", "Los Angeles HD")) %>%
  select(county_lhj, year, population)


# Append and save -------------------------------------------------

lhjJuly_final <- bind_rows(
  rename(countyEstimates, county_lhj = county),
  select(cityEstimates, county_lhj = place, year, population = city_population),
  hdEstimates
  ) %>% 
  arrange(county_lhj, year)

# Data Quality checks
if (FALSE) {
  colSums(is.na(lhjJuly_final)) # Should return 0s, meaning there are no missing values
  # Berkeley to Alameda HD = Alameda County
  al <- lhjJuly_final %>% 
    filter(county_lhj %in% c("Berkeley", "Alameda HD", "Alameda")) %>% 
    pivot_wider(names_from = county_lhj, values_from = population) %>% 
    mutate(eq = Berkeley + `Alameda HD` == Alameda, 
           city_county_prop = Berkeley / Alameda)
}


saveRDS(lhjJuly_final, paste0(outputDir, "lhj-population-total.RDS"))
write_csv(lhjJuly_final, paste0(outputDir, "lhj-population-total.csv"))

```
