---
title: "Create Total LHJ Population Data"
author: "Jaspreet Kang"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, eval = FALSE)
```

# Introduction

This RMarkdown file creates a dataset with total July population estimates. This dataset is used in *3-Create-LHJ-Population-ARS.Rmd* to perform the optional step of adjusting the DOF Jan 1st P3 estimates to July 1st estimates. Therefore, this Rmarkdown file must be executed only if the adjustment of DOF P3 data is needed or desired.

## Population Data Required

Four outputs are used from *1a-Process-Historical-Raw-Data.Rmd* and *1b-Process-Current-Raw-Data.Rmd*:

1.  *intermediatePopData/dof-e4-processed-historical.RDS*

    1.  Contains city to county proportions (using January 1st estimates) from 1970-2010.

2.  *intermediatePopData/dof-e4-processed-current.RDS*

    1.  Contains city to county proportions (using January 1st estimates) from 2011-2023.

3.  *intermediatePopData/dof-e6-processed-historical.RDS*

    1.  Processed DOF July 1st estimates for counties from 1947-2009

4.  *intermediatePopData/dof-e6-processed-current.RDS*

    1.  Processed DOF July 1st estimates for counties from 2010-2023

## Method

To calculate total July population estimates for the 3 cities, multiply each city-to-county proportion to its respective county July population to get a city July population estimate.

Then, to calculate total July population estimates for Alameda and Los Angeles Health Departments:

1.  Alameda HD July population = Alameda County July population - Berkeley City July population

2.  Los Angeles HD July population = Los Angeles County July population - (Long Beach City July population + Pasadena City July population)

## Data File Outputs

Output is saved as a csv and RDS file into *finalPopData/*:

1.  *lhj-population-total.RDS*
2.  *lhj-population-total.csv*

# Setup

```{r}


# Disable scientific notation
options(scipen=999)

## Load packages ----------------------------------------------
library(readxl) # Read in excel files
library(readr) # Read in and write csv files
library(dplyr) # Data manipulation
library(tidyr) # Data reshaping

# City LHJs
cities <- c("Berkeley", "Long Beach", "Pasadena")

# Directory to read input data files from
inputDir <- "intermediatePopData/"

# Directory to save output files to
outputDir <- "finalPopData/"

```

# Read in data

Read in the processed DOF E-4 and E-6 data - both historical and current.

```{r}

e4_historical <- readRDS(paste0(inputDir, "dof-e4-processed-historical.RDS"))
e4_current <- readRDS(paste0(inputDir, "dof-e4-processed-current.RDS"))
e4 <- bind_rows(e4_historical, e4_current) # Combine e4 historical and current


e6_historical <- readRDS(paste0(inputDir, "dof-e6-processed-historical.RDS"))
e6_current <- readRDS(paste0(inputDir, "dof-e6-processed-current.RDS"))
e6 <- bind_rows(e6_historical, e6_current) # Combine e6 historical and current

```

# Calculate LHJ Total Estimates

```{r}

# County Estimates ---------------------------------------

countyEstimates <- e6

# City Estimates -----------------------------------------

cityEstimates <- e4 %>% 
  left_join(rename(e6, county_population = population)) %>% 
  mutate(city_population = city_county_proportion * county_population)


# Health Department Estimates ----------------------------
  
hdEstimates <- cityEstimates %>%
  mutate(place = ifelse(place == "Berkeley", "Berkeley", "Pasadena/Long Beach")) %>% 
  group_by(place, year, county) %>% 
  summarise(county_population = mean(county_population), 
            city_population = sum(city_population)) %>% 
  ungroup() %>% 
  mutate(population = county_population - city_population, 
         county_lhj = ifelse(place == "Berkeley", "Alameda HD", "Los Angeles HD")) %>%
  select(county_lhj, year, population)


# Append and save -------------------------------------------------

lhjJuly_final <- bind_rows(
  rename(countyEstimates, county_lhj = county),
  select(cityEstimates, county_lhj = place, year, population = city_population),
  hdEstimates
  ) %>% 
  arrange(county_lhj, year)

# Data Quality checks
if (FALSE) {
  colSums(is.na(lhjJuly_final)) # Should return 0s, meaning there are no missing values
}


saveRDS(lhjJuly_final, paste0(outputDir, "lhj-population-total.RDS"))
write_csv(lhjJuly_final, paste0(outputDir, "lhj-population-total.csv"))

```
