---
title: "Create LHJ Sex-Age-Race/Ethnicity Population Data"
author: "CDPH/LHJ Population Task Force"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, eval = FALSE)
```

# Introduction

This RMarkdown file creates two datasets with age-race/ethnicity-sex population estimates for:

-   All 58 counties
-   The 3 city-based jurisdictions: Berkeley, Long Beach, and Pasadena
-   Alameda Health Department
-   Los Angeles Health Department

The first dataset contains estimates from 2000-2023, and is recommended to use for trends.

The second dataset contains estimates from 2020-2023, and is recommended to use when reporting data on recent years only.

## Population Data Required

Outputs from *1a-Process-Historical-Raw-Data.Rmd* and *1b-Process-Current-Raw-Data.Rmd*:

1.  intermediatePopData/census-city-ars-historical.RDS & intermediatePopData/census-city-ars-current.RDS

    1.  City-level Sex-Age-Race/Ethnicity population estimates pulled from the 2000, 2010, and 2020 Decennial.

2.  intermediatePopData/census-county-ars-historical.RDS & intermediatePopData/census-county-ars-current.RDS

    1.  County-level Sex-Age-Race/Ethnicity population estimates pulled from the 2000, 2010, and 2020 Decennial.

3.  intermediatePopData/dof-p3-processed-historical.RDS & intermediatePopData/dof-p3-processed-current.RDS

    1.  Processed Sex-Age-Race/Ethnicity population estimates from 2000-2023

4.  intermediatePopData/lhj-population-total.RDS

    1.  Total July LHJ population estimates

Final outputs are saved as csv and RDS files in finalPopData/:

1.  lhj-pop-ars-2000-cy
2.  lhj-pop-ars-2020-cy

Note: 'cy' means 'current year' which is 2023.

# Setup

```{r}


# Disable scientific notation
options(scipen=999)

## Load packages ----------------------------------------------
library(readxl) # Read in excel files
library(readr) # Read in and write csv files
library(dplyr) # Data manipulation
library(tidyr) # Data reshaping

# Cities
cities <- c("Berkeley", "Long Beach", "Pasadena")

# Directories
inputDir <- "intermediatePopData/"
outputDir <- "finalPopData/"
```

# Data Processing

## Read in data

```{r}

# Census city data
censusCity_historical <- readRDS(paste0(inputDir, "census-city-ars-historical.RDS"))
censusCity_current <- readRDS(paste0(inputDir, "census-city-ars-current.RDS"))
censusCity <- bind_rows(censusCity_historical, censusCity_current)

# Census county data
censusCounty_historical <- readRDS(paste0(inputDir, "census-county-ars-historical.RDS"))
censusCounty_current <- readRDS(paste0(inputDir, "census-county-ars-current.RDS"))
censusCounty <- bind_rows(censusCounty_historical, censusCounty_current)

# DOF P3 data
p3_historical <- readRDS(paste0(inputDir, "dof-p3-processed-historical.RDS")) %>% mutate(source = "old dof")
p3_current <- readRDS(paste0(inputDir, "dof-p3-processed-current.RDS"))
p3 <- bind_rows(p3_historical, p3_current)

# July 1 city, HD, and county totals - for adjustment
lhjTotal <- readRDS(paste0(inputDir, "lhj-population-total.RDS"))



```

## Perform Adjustment

Adjust **DOF-P3** county sex-age-race/ethnicity population estimates to the "better" county July 1st estimates (in the "lhj-population-total" file). Note that the county populations in "lhj-population-total" are identical to the county populations in **DOF-E6**

Adjustment formula:

-   (DOF-P3 County ARS Strata Population) / (DOF-P3 Total County Population) \* (DOF-E6 July 1st Total County Population)

Note: The total county population estimates in **DOF-P3** already match the July county estimates in **DOF-E6** for 2000-2009.

```{r}

countyARS <- p3 %>% 
    rename(popP3 = population)%>% 
    group_by(county, year, source)%>% 
    mutate(county_population = sum(popP3)) %>% 
    ungroup() %>% 
    mutate(ratioStrata = popP3 / county_population) %>% 
    left_join(rename(lhjTotal, popJuly = population), by = c("county" = "county_lhj", "year")) %>% 
    mutate(population = ratioStrata * popJuly) %>% 
    select(county_lhj = county, year, sex, raceEth, age, source, population)
  

```

## Redistribute "Other" Race

Redistribute 'Other' Race (Non-Hispanic) in Census City and County Data

Process:

1.  Within each year-geography-sex-age strata:

    1.  Calculate the total (Non-Hispanic) population excluding "Other"

        1.  popTotal = popAIAN + popAsian + popBlack + popNHPI + popMultiRace + popWhite

    2.  Calculate each race/ethnicity-specific population proportion

        1.  propAIAN = popAIAN/popTotal

        2.  propAsian = popAsian/popTotal

        3.  etc

    3.  Multiply each race/ethnicity-specific population proportion to "Other" Race population

        1.  popOther_AIAN = propAIAN \* popOther

        2.  popOther_Asian= propAsian \* popOther

        3.  etc

    4.  For each race/ethnicity, add the value obtained from prior step to the original population estimate.

        1.  popAIAN_new = popOther_AIAN + popAIAN

        2.  popAsian_new = popOther_Asian + popAsian

```{r}

# Redistribute 'Other' Race ----------------------------------------------------------------------------
cityOther <- censusCity %>% 
  filter(raceEth == "Other") %>% 
  rename(raceOther = population) %>% 
  select(-raceEth)

cityLatino <- censusCity %>% 
  filter(raceEth == "Latino")

cityTotal <- censusCity %>% 
  filter(raceEth == "Total")

cityRace <- censusCity %>% 
  filter(!raceEth %in% c("Other", "Latino", "Total")) %>% 
  group_by(year, GEOID, NAME, sex, age) %>% 
  mutate(raceTotal = sum(population)) %>% 
  ungroup() %>% 
  mutate(raceProp = population / raceTotal) %>% 
  left_join(cityOther) %>% 
  mutate(populationNew = population + (raceOther * raceProp))

censusCity <- cityRace %>% 
  select(year, GEOID, NAME, sex, age, raceEth, population = populationNew) %>% 
  mutate(population = ifelse(is.nan(population), 0, population)) %>% 
  bind_rows(cityLatino)



countyOther <- censusCounty %>% 
  filter(raceEth == "Other") %>% 
  rename(raceOther = population) %>% 
  select(-raceEth)

countyLatino <- censusCounty %>% 
  filter(raceEth == "Latino")

countyTotal <- censusCounty %>% 
  filter(raceEth == "Total")

countyRace <- censusCounty %>% 
  filter(!raceEth %in% c("Other", "Latino", "Total")) %>% 
  group_by(year, GEOID, NAME, sex, age) %>% 
  mutate(raceTotal = sum(population)) %>% 
  ungroup() %>% 
  mutate(raceProp = population / raceTotal) %>% 
  left_join(countyOther) %>% 
  mutate(populationNew = population + (raceOther * raceProp))

censusCounty <- countyRace %>% 
  select(year, GEOID, NAME, sex, age, raceEth, population = populationNew) %>% 
  mutate(population = ifelse(is.nan(population), 0, population)) %>% 
  bind_rows(countyLatino)


# Check Totals match:
checkCounty <- censusCounty %>% 
  filter(sex == "Total", age == "Total") %>% 
  group_by(year, GEOID, NAME) %>% 
  summarise(checkPop = sum(population)) %>% 
  full_join(countyTotal) %>% 
  mutate(diff = checkPop - population)

checkCity <- censusCity %>% 
  filter(sex == "Total", age == "Total") %>% 
  group_by(year, GEOID, NAME) %>% 
  summarise(checkPop = sum(population)) %>% 
  full_join(cityTotal) %>% 
  mutate(diff = checkPop - population)

```

## Further Processing of Census Data

1.  Further process the Census city and county-level data
2.  Merge Census city and county-level ARS data. Then calculate the proportion of the city-level ARS strata to the population of the corresponding county-level ARS strata.

```{r}

# Further processing ----------------------------------
censusCityARS <- censusCity %>%  
  filter(NAME %in% c("Berkeley city, California", "Pasadena city, California", "Long Beach city, California"), 
         raceEth != "Total", sex != "Total", age != "Total") %>% 
  mutate(NAME = sub(" city, California", "", NAME), 
         county = ifelse(NAME == "Berkeley", "Alameda", "Los Angeles") 
         ) %>% 
  group_by(year, NAME, county, sex, age, raceEth) %>% 
  summarise(populationCity = sum(population)) %>% 
  ungroup()

censusCountyARS <- censusCounty %>% 
  filter(NAME %in% c("Alameda County, California", "Los Angeles County, California"), 
         raceEth != "Total", sex != "Total", age != "Total") %>% 
  mutate(NAME = sub(" County, California", "", NAME) 
         ) %>% 
  group_by(year, county = NAME, sex, age, raceEth) %>% 
  summarise(populationCounty = sum(population)) %>% 
  ungroup()

# Merge Census City and County data frame, and calculate city to county proportion
censusCityCountyARS <- censusCityARS %>% 
  left_join(censusCountyARS, by = c("year", "county", "sex", "age", "raceEth")) %>% 
  mutate(cityCountyProp = populationCity / populationCounty, 
         cityCountyProp = ifelse(is.nan(cityCountyProp), 0, cityCountyProp)) 
```

## Calculate city-level sex-age-race/ethnicity population

Multiply the Census Decennial sex-age-race/ethnicity strata's city:county proportion to the corresponding **DOF-P3** (adjusted to DOF-E6) county-level sex-age-race/ethnicity strata's population estimate

Note: Census Decennial data only has data for 2000, 2010, and 2020. Therefore:

-   decennial 2000 city:county proportions are multiplied to DOF P3 2000-2009 data
-   decennial 2010 city:county proportions are multiplied to DOF P3 2010-2019 data
-   decennial 2020 city:county proportions are multiplied to DOF P3 2020-2023 data

```{r}

joinCityCounty <- function(myYear, mySource) {
  myYears <- myYear:(myYear+9)
  
  tCity <- censusCityCountyARS %>% filter(year == myYear) %>% select(-year)
  
  if (mySource == "old dof") {
    tCounty <- countyARS %>% filter(year %in% myYears, source == mySource)
  } else {
    tCounty <- countyARS %>% filter(year %in% myYears, source == mySource)
  }
  
  
  tCity %>% 
    left_join(tCounty, by = c("county" = "county_lhj", "sex", "age", "raceEth"), relationship = "many-to-many") 
}

# Trends data: Use pre-2020 P3 
cityARS_trends_final <- bind_rows(
  joinCityCounty(2000, "old dof"),
  joinCityCounty(2010, "old dof"),
  joinCityCounty(2020, "old dof")
) %>% 
  mutate(population = cityCountyProp * population)%>%
  select(county_lhj = NAME, year, sex, age, raceEth, source, population)

# Recent data: Use post-2020 P3
cityARS_recent_final <- joinCityCounty(2020, "new dof") %>% 
  mutate(population = cityCountyProp * population) %>%
  select(county_lhj = NAME, year, sex, age, raceEth, source, population)

cityARS_final <- bind_rows(cityARS_trends_final, cityARS_recent_final)

# Adjust to DOF city
cityARS_final_adj <- cityARS_final %>% 
  group_by(county_lhj, year, source) %>% 
  mutate(cityPopCalc = sum(population)) %>% 
  ungroup() %>% 
  left_join(rename(lhjTotal, cityPopDOF = population)) %>% 
  mutate(popAdj = (cityPopDOF / cityPopCalc) * population) %>% 
  select(county_lhj, year, sex, age, raceEth, source, population = popAdj)
  


# Data Quality checks
if (FALSE) {
  
  # Check if LHJ totals match
  checkTotals <- cityARS_final_adj %>% 
    group_by(county_lhj, year, source) %>% 
    summarise(cityPopCalc = sum(population)) %>%
    ungroup() %>% 
    left_join(lhjTotal) %>% 
    mutate(diff = cityPopCalc - population)
  
table(cityARS_final_adj$county_lhj, cityARS_final_adj$year, useNA = "ifany")
length(unique(cityARS_final_adj$sex)) * length(unique(cityARS_final_adj$age)) * length(unique(cityARS_final_adj$raceEth))
}


```

## Calculate health department-level sex-age-race/ethnicity population

Calculate sex-age-race/ethnicity estimates at the health department level:

-   Alameda HD ARS Population = (Alameda County ARS Population) - (Berkeley ARS Population)

-   Los Angeles HD ARS Population = (Los Angeles County ARS Population) - (Long Beach ARS Population + Pasadena ARS Population)

Note: Negative population estimates are calculated for around 34 sex-age-race-health department stratas. These all occur in NH/PI, Multi-Race, or AI/AN 95+ year olds whose populations are very small. These negative values are set to 0.

```{r}

# Health Department estimates ----------------------
hdARS <- cityARS_final_adj %>% 
  mutate(county_lhj = ifelse(county_lhj == "Berkeley", "Berkeley", "Pasadena/Long Beach"))%>% 
  group_by(city = county_lhj, year, sex, age, raceEth, source) %>% 
  summarise(city_population = sum(population))%>% 
  ungroup() %>% 
  mutate(county_lhj = ifelse(city == "Berkeley", "Alameda", "Los Angeles")) %>% 
  left_join(rename(countyARS, county_population = population))%>% 
  mutate(city_population = ifelse(city_population > county_population, county_population, city_population),
         population = county_population - city_population,
         county_lhj = ifelse(city == "Berkeley", "Alameda HD", "Los Angeles HD"))%>% 
  select(county_lhj, year, sex, age, raceEth, source, population)


```

## Combine and Save Data

Combine city, health department, and county level data; Save data

```{r}

# Append and save --------------------------------
lhjARS <- bind_rows(countyARS, cityARS_final_adj, hdARS) %>% 
  select(county_lhj, year, sex, age, raceEth, source, population) %>% 
  mutate(age = ifelse(age == "100+", "100", age), 
         age = as.numeric(age)) %>% 
  arrange(county_lhj, year, sex, age, raceEth) %>% 
  rename(race_eth = raceEth)

# lhjARS_csv <- lhjARS %>% # Temporary: 2000- csv too large to push to github
#   filter(year >= 2010)

# Checks -----------------------------------------
if (FALSE) {
  colSums(is.na(lhjARS))
}

# Separate
lhjARS_trends <- lhjARS %>% 
  filter(source == "old dof") %>% 
  select(-source)

lhjARS_recent <- lhjARS %>% 
  filter(source == "new dof") %>% 
  select(-source)

# Check
table(lhjARS_trends$year, useNA = "ifany")
table(lhjARS_recent$year, useNA = "ifany")

check <- lhjARS_trends %>% 
  filter(county_lhj %in% c("Berkeley", "Alameda", "Alameda HD")) %>% 
  pivot_wider(names_from = county_lhj, values_from = population) %>% 
  mutate(newAlameda = Berkeley + `Alameda HD`,
         diff = abs(Alameda - newAlameda),
         check = abs(Alameda - newAlameda) < 1e-5)

checkTotals <- lhjARS %>% 
  filter(!grepl("HD|Berkeley|Pasadena|Beach", county_lhj))%>% 
  bind_rows(mutate(., race_eth = "Total")) %>% 
  group_by(year, source, race_eth) %>% 
  summarise(population = sum(population)) %>% 
  ungroup()

library(ggplot2)
ggplot(checkTotals, aes(x = year, y = population, color = source)) +
  geom_line() +
  facet_wrap(vars(race_eth), scales = "free_y") +
  scale_x_continuous(minor_breaks = 2000:2023, breaks = 2000:2023, labels = 2000:2023) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))

# Save data -------------------------------------


saveRDS(lhjARS_trends, paste0(outputDir, "lhj-pop-ars-2000-cy.RDS"))
write_csv(lhjARS_trends, paste0(outputDir, "lhj-pop-ars-2000-cy.csv"))
saveRDS(lhjARS_recent, paste0(outputDir, "lhj-pop-ars-2020-cy.RDS"))
write_csv(lhjARS_recent, paste0(outputDir, "lhj-pop-ars-2020-cy.csv"))


```
