---
title: "Create LHJ Sex-Age-Race/Ethnicity Population Data"
author: "Jaspreet Kang"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, eval = FALSE)
```

# Introduction

This RMarkdown file creates two datasets with age-race/ethnicity-sex population estimates for:

-   All 58 counties
-   The 3 city-based jurisdictions: Berkeley, Long Beach, and Pasadena
-   Alameda Health Department
-   Los Angeles Health Department

The first dataset contains January 1st estimates while the second dataset contains adjusted July 1st estimates.

## Population Data Required

Outputs from *1a-Process-Historical-Raw-Data.Rmd* and *1b-Process-Current-Raw-Data.Rmd*:

1.  intermediatePopData/census-city-ars-historical.RDS & intermediatePopData/census-city-ars-current.RDS

    1.  City-level Sex-Age-Race/Ethnicity population estimates pulled from the 2000, 2010, and 2020 Decennial.

2.  intermediatePopData/census-county-ars-historical.RDS & intermediatePopData/census-county-ars-current.RDS

    1.  County-level Sex-Age-Race/Ethnicity population estimates pulled from the 2000, 2010, and 2020 Decennial.

3.  intermediatePopData/dof-p3-processed-historical.RDS & intermediatePopData/dof-p3-processed-current.RDS

    1.  Processed Sex-Age-Race/Ethnicity population estimates from 2000-2023

(Optional) One output is used from 2-Create-LHJ-Population-Total.Rmd

1.  finalPopData/lhj-population-total.RDS

    1.  Total July population estimates

Final outputs are saved as csv and RDS files in finalPopData/:

1.  lhj-population-ars
2.  lhj-population-ars-adjJuly

# Setup

Note: Set the object **adjust** below to TRUE if final LHJ ARS data should be adjusted from January estimates to July estimates; Otherwise set **adjust** to FALSE

```{r}
adjust <- FALSE

# Disable scientific notation
options(scipen=999)

## Load packages ----------------------------------------------
library(readxl) # Read in excel files
library(readr) # Read in and write csv files
library(dplyr) # Data manipulation
library(tidyr) # Data reshaping

# Cities
cities <- c("Berkeley", "Long Beach", "Pasadena")

# Directories
inputDir <- "intermediatePopData/"
outputDir <- "finalPopData/"
```

# Data Processing

## Read in data

```{r}

# Census city data
censusCity_historical <- readRDS(paste0(inputDir, "census-city-ars-historical.RDS"))
censusCity_current <- readRDS(paste0(inputDir, "census-city-ars-current.RDS"))
censusCity <- bind_rows(censusCity_historical, censusCity_current)

# Census county data
censusCounty_historical <- readRDS(paste0(inputDir, "census-county-ars-historical.RDS"))
censusCounty_current <- readRDS(paste0(inputDir, "census-county-ars-current.RDS"))
censusCounty <- bind_rows(censusCounty_historical, censusCounty_current)

# DOF P3 data
p3_historical <- readRDS(paste0(inputDir, "dof-p3-processed-historical.RDS"))
p3_current <- readRDS(paste0(inputDir, "dof-p3-processed-current.RDS"))
p3 <- bind_rows(p3_historical, p3_current)

# July 1 city, HD, and county totals - for optional adjustment
if (adjust) {
  lhjTotal <- readRDS(paste0(outputDir, "lhj-population-total.RDS"))
}



```

## (Optional) Perform Adjustment

If the **adjust** object above is set to TRUE, then adjust DOF P3 Jan 1st sex-age-race/ethnicity population estimates to July 1st estimates.

Adjustment formula:

-   (DOF P3 Jan 1st County ARS Strata Population) / (DOF P3 Jan 1st Total County Population) \* (DOF E6 July 1st Total County Population)

Note: The total county population estimates in DOF P3 should already match the latest July county estimates in DOF E6. This is true for 2000-2009 & 2020-2023. However for 2010-2019, the total county population estimates are slightly different due to the several revisions DOF made to their P3 Vintage 2020 (pre Decennial 2020) file.

```{r}

if (adjust) {
  countyARS <- p3 %>% 
    rename(popJan = population)%>% 
    group_by(county, year)%>% 
    mutate(county_population = sum(popJan)) %>% 
    ungroup() %>% 
    mutate(ratioStrata = popJan / county_population) %>% 
    left_join(rename(lhjTotal, popJuly = population), by = c("county" = "county_lhj", "year")) %>% 
    mutate(population = ratioStrata * popJuly) %>% 
    select(county_lhj = county, year, sex, raceEth, age, population)
} else {
  countyARS <- p3 %>% 
    rename(county_lhj = county)
}
  

```

## Redistribute "Other" Race

Redistribute 'Other' Race in Census City and County Data

Process:

1.  Within each year-geography-sex-age strata:

    1.  Calculate the total population

        1.  popTotal = popAIAN + popAsian + popBlack + popLatino + popNHPI + popMultiRace + popWhite

            1.  Or simply: popTotal = popTotal - popOther

    2.  Calculate each race/ethnicity-specific population proportion

        1.  propAIAN = popAIAN/popTotal

        2.  propAsian = popAsian/popTotal

        3.  etc

    3.  Multiply each race/ethnicity-specific population proportion to "Other" Race population

        1.  popOther_AIAN = propAIAN \* popOther

        2.  popOther_Asian= propAsian \* popOther

        3.  etc

    4.  For each race/ethnicity, add the value obtained from prior step to the original population estimate.

        1.  popAIAN_new = popOther_AIAN + popAIAN

        2.  popAsian_new = popOther_Asian + popAsian

```{r}

# Redistribute 'Other' Race ----------------------------------------------------------------------------
cityOther <- censusCity %>% 
  filter(raceEth == "Other") %>% 
  rename(raceOther = population) %>% 
  select(-raceEth)

cityRace <- censusCity %>% 
  filter(raceEth != "Other") %>% 
  group_by(year, GEOID, NAME, sex, age) %>% 
  mutate(raceTotal = sum(population)) %>% 
  ungroup() %>% 
  mutate(raceProp = population / raceTotal) %>% 
  left_join(cityOther) %>% 
  mutate(populationNew = population + (raceOther * raceProp))

censusCity <- cityRace %>% 
  select(year, GEOID, NAME, sex, age, raceEth, population = populationNew) %>% 
  mutate(population = ifelse(is.nan(population), 0, population))


countyOther <- censusCounty %>% 
  filter(raceEth == "Other") %>% 
  rename(raceOther = population) %>% 
  select(-raceEth)

countyRace <- censusCounty %>% 
  filter(raceEth != "Other") %>% 
  group_by(year, GEOID, NAME, sex, age) %>% 
  mutate(raceTotal = sum(population)) %>% 
  ungroup() %>% 
  mutate(raceProp = population / raceTotal) %>% 
  left_join(countyOther) %>% 
  mutate(populationNew = population + (raceOther * raceProp))

censusCounty <- countyRace %>% 
  select(year, GEOID, NAME, sex, age, raceEth, population = populationNew) %>% 
  mutate(population = ifelse(is.nan(population), 0, population))

```

## Further Processing of Census Data

1.  Further process the Census city and county-level data
2.  Merge Census city and county-level ARS data. Then calculate the proportion of the city-level ARS strata to the population of the corresponding county-level ARS strata.

```{r}

# Further processing ----------------------------------
censusCityARS <- censusCity %>%  
  filter(NAME %in% c("Berkeley city, California", "Pasadena city, California", "Long Beach city, California"), 
         raceEth != "Total", sex != "Total", age != "Total") %>% 
  mutate(NAME = sub(" city, California", "", NAME), 
         county = ifelse(NAME == "Berkeley", "Alameda", "Los Angeles") 
         ) %>% 
  group_by(year, NAME, county, sex, age, raceEth) %>% 
  summarise(populationCity = sum(population)) %>% 
  ungroup()

censusCountyARS <- censusCounty %>% 
  filter(NAME %in% c("Alameda County, California", "Los Angeles County, California"), 
         raceEth != "Total", sex != "Total", age != "Total") %>% 
  mutate(NAME = sub(" County, California", "", NAME) 
         ) %>% 
  group_by(year, county = NAME, sex, age, raceEth) %>% 
  summarise(populationCounty = sum(population)) %>% 
  ungroup()

# Merge Census City and County data frame, and calculate city to county proportion
censusCityCountyARS <- censusCityARS %>% 
  left_join(censusCountyARS, by = c("year", "county", "sex", "age", "raceEth")) %>% 
  mutate(cityCountyProp = populationCity / populationCounty, 
         cityCountyProp = ifelse(is.nan(cityCountyProp), 0, cityCountyProp)) 
```

## Calculate city-level sex-age-race/ethnicity population

Multiply the Census Decennial sex-age-race/ethnicity strata's city:county proportion to the corresponding DOF P3 county-level sex-age-race/ethnicity strata's population estimate (either July estimate if adjustment was performed; Otherwise the original January estimate)

Note: Census Decennial data only has data for 2000, 2010, and 2020. Therefore:

-   decennial 2000 city:county proportions are multiplied to DOF P3 2000-2009 data
-   decennial 2010 city:county proportions are multiplied to DOF P3 2010-2019 data
-   decennial 2020 city:county proportions are multiplied to DOF P3 2020-2023 data

```{r}

joinCityCounty <- function(myYear) {
  myYears <- myYear:(myYear+9)
  
  tCity <- censusCityCountyARS %>% filter(year == myYear) %>% select(-year)
  tCounty <- countyARS %>% filter(year %in% myYears)
  tCity %>% 
    left_join(tCounty, by = c("county" = "county_lhj", "sex", "age", "raceEth"), relationship = "many-to-many") 
}

cityARS_final <- bind_rows(
  joinCityCounty(2000),
  joinCityCounty(2010),
  joinCityCounty(2020)
) %>% 
  mutate(population = cityCountyProp * population) %>%
  select(county_lhj = NAME, year, sex, age, raceEth, population)

# Data Quality checks
if (FALSE) {
  table(cityARS_final$county_lhj, cityARS_final$year, useNA = "ifany")
length(unique(cityARS_final$sex)) * length(unique(cityARS_final$age)) * length(unique(cityARS_final$raceEth))
}


```

## Calculate health department-level sex-age-race/ethnicity population

Calculate sex-age-race/ethnicity estimates at the health department level:

-   Alameda HD ARS Population = (Alameda County ARS Population) - (Berkeley ARS Population)

-   Los Angeles HD ARS Population = (Los Angeles County ARS Population) - (Long Beach ARS Population + Pasadena ARS Population)

```{r}

# Health Department estimates ----------------------
hdARS <- cityARS_final %>% 
  mutate(county_lhj = ifelse(county_lhj == "Berkeley", "Berkeley", "Pasadena/Long Beach"))%>% 
  group_by(city = county_lhj, year, sex, age, raceEth) %>% 
  summarise(city_population = sum(population))%>% 
  ungroup() %>% 
  mutate(county_lhj = ifelse(city == "Berkeley", "Alameda", "Los Angeles")) %>% 
  left_join(rename(countyARS, county_population = population))%>% 
  mutate(population = county_population - city_population,
         county_lhj = ifelse(city == "Berkeley", "Alameda HD", "Los Angeles HD"))%>% 
  select(county_lhj, year, sex, age, raceEth, population)


```

## Combine and Save Data

Combine city, health department, and county level data; Save data

```{r}

# Append and save --------------------------------
lhjARS <- bind_rows(countyARS, cityARS_final, hdARS) %>% 
  select(county_lhj, year, sex, age, raceEth, population) %>% 
  arrange(county_lhj, year, sex, age, raceEth)

# Checks -----------------------------------------
if (FALSE) {
  colSums(is.na(lhjARS))
}


# Save data -------------------------------------

if (adjust) {
  saveRDS(lhjARS, paste0(outputDir, "lhj-population-ars-adjJuly.RDS"))
  write_csv(lhjARS, paste0(outputDir, "lhj-population-ars-adjJuly.csv"))
} else {
  saveRDS(lhjARS, paste0(outputDir, "lhj-population-ars.RDS"))
  write_csv(lhjARS, paste0(outputDir, "lhj-population-ars.csv"))
}

```
