---
title: "Investigating DOF 'Estimates' Files"
author: "Jaspreet Kang"
date: "3/27/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(readxl)
library(readr)
library(dplyr)
library(tidyr)
library(janitor)
library(lubridate)

# Info Files
countyLink <- read_xlsx("Standards/countyLink.xlsx") %>% 
  select(county = countyName)

```

# Read in DOF 'Estimates' data

```{r}

# E1 File
e1_2022 <- read_xlsx("E1/E-1_2022_InternetVersion.xlsx", sheet = "E-1 CountyState2022", skip = 4)

# E4 Files
e4_70_80 <- read_xlsx("E4/E_4-1970-1980.xlsx", skip = 5, col_names = c("Place", "4/1/1970", paste0("1/1/", 1971:1980), "4/1/1980"))

e4_80_90 <- read_xlsx("E4/90E-4.xlsx", skip = 93, col_names = c("County", paste0("1/1/", 1981:1990)))


e4_90_00 <- read_xlsx("E4/E-4_90-00_Rpt.xlsx", sheet = "Table 2  City Pop Est", skip = 4, col_names = c("County", "4/1/1990", paste0("1/1/", 1991:2000), "4/1/2000"))

e4_00_10 <- read_xlsx("E4/E4_2000-2010_Report_Final_EOC_000.xlsx", sheet = "Table 2 City County", col_names = c("County", "4/1/2000", paste0("1/1/", 2001:2010), "4/1/2010"), skip = 4)

e4_10_20 <- read_xlsx("E4/E-4_2010-2020-Internet-Version.xlsx", sheet = "Table 2 City County", skip = 2, col_names = c("County", "4/1/2010", paste0("1/1/", 2011:2020)))

e4_20_22 <- read_xlsx("E4/E-4_2022_InternetVersion.xlsx", sheet = "Table 2 City County", skip = 2, col_names = c("County", "4/1/2020", "1/1/2021", "1/1/2022"))

# E6 Files
e6_47_69 <- read_xlsx("E6/E-6CntyPopEstmts1947-1969.xlsx", skip = 3, 
                      col_names = c("County", 
                                    "4/1/1940",
                                    paste0("7/1/", 1947:1949),
                                    "4/1/1950", 
                                    paste0("7/1/", 1950:1959), 
                                    "4/1/1960", 
                                    paste0("7/1/", 1960:1969), 
                                    "4/1/1970"))


e6_70_80 <- read_xlsx("E6/E-6_70-90final.xlsx", sheet = "E-6 70-80", skip = 2) %>% 
  select(County = 1, Year, july1_population = 3)

e6_80_90 <- read_xlsx("E6/E-6_70-90final.xlsx", sheet = "E-6 80-90", skip = 2) %>% 
  select(County = 1, Year, july1_population = 3)


e6_90_00 <- read_xlsx("E6/E-6_90-00.xlsx", skip = 2) %>% 
  select(County = 1, Year = 2, july1_population = 3)

e6_00_10 <- read_xlsx("E6/E-6_Report_July_2000-2010_updated_with_2010_census.xlsx", skip = 3) %>% 
  select(County = 1, Year = 2, july1_population = 3)
  
e6_10_21 <- read_xlsx("E6/E-6_Report_July_2010-2021_w.xlsx", sheet = "E-6 2010-2021 Report", skip = 3) %>% 
  select(County = 1, Year = 2, july1_population = 3)

e6_20_22 <- read_xlsx("E6/E-6_Report_July_2020-2022_w.xlsx", sheet = "E-6 2020-2022 Report", skip = 3) %>% 
  select(County = 1, Year = 2, july1_population = 3)


```

### Pasadena and South Pasadena appear in every E4 file. 

# Clean Up Data

```{r}

# Data Processing ------------------------------------------

## E4 - 1970-1980 ---------
counties <- paste0(countyLink$county, " County")

e4_70_80 <- lapply(counties, function(x) {
  
  tDat <- e4_70_80 %>% 
    filter(grepl("County", Place))
  
  countyIndex <- which(tDat$Place == x)
  tDat[countyIndex, 2:ncol(tDat)] <- tDat[countyIndex + 1, 2:ncol(tDat)]
  
  tDat %>% 
    filter(Place == x)
  
}) %>% 
  bind_rows() %>% 
  rename(County = Place) %>% 
  mutate(County = sub("*.County", "", County), 
         across(.cols = -1, .fns = as.numeric)) %>% 
  adorn_totals("row", name = "CALIFORNIA")
  
## E4 - 1980-1990 ---------

caIndex <- min(which(e4_80_90$County == "CALIFORNIA"))
e4_80_90 <- e4_80_90[1:caIndex, ] %>% 
  filter(!is.na(County)) %>% 
  mutate(across(.cols = -1, .fns = as.numeric))

## E4 - 1990-2000 ---------

e4_90_00 <- e4_90_00 %>% 
  filter(!is.na(County)) %>% 
  mutate(County = ifelse(County == "State Total", "CALIFORNIA", County))


## E4 - 2000-2010 ---------

e4_00_10 <- e4_00_10 %>% 
  filter(!is.na(County)) %>% 
  mutate(County = ifelse(County == "State Total", "CALIFORNIA", County))


## E4 - 2010-2020 ---------

e4_10_20 <- e4_10_20 %>% 
  filter(!is.na(County)) %>% 
  mutate(County = ifelse(County == "State Total", "CALIFORNIA", County))


## E4 - 2020-2022 ---------

e4_20_22 <- e4_20_22 %>% 
  filter(!is.na(County)) %>% 
  mutate(County = ifelse(County == "State Total", "CALIFORNIA", County))

# Check for duplicate dates --------
data.frame(date = c(names(e4_70_80), 
           names(e4_80_90), 
           names(e4_90_00), 
           names(e4_00_10), 
           names(e4_10_20), 
           names(e4_20_22)), stringsAsFactors = F) %>% 
  count(date) %>% 
  filter(n > 1)

identical(e4_90_00$`4/1/2000`, e4_00_10$`4/1/2000`)
identical(e4_00_10$`4/1/2010`, e4_10_20$`4/1/2010`)

# E4 - Final ------------

e4_wide <- e4_70_80 %>% 
  full_join(e4_80_90) %>% 
  full_join(select(e4_90_00, -`4/1/2000`)) %>% 
  full_join(select(e4_00_10, -`4/1/2010`)) %>% 
  full_join(e4_10_20) %>% 
  full_join(e4_20_22)

e4_long <- e4_wide %>% 
  pivot_longer(-County, names_to = "Date", values_to = "Population") %>% 
  mutate(Date = lubridate::mdy(Date))

```

```{r}

# Data Processing Continued ---------------

## E6 - 1947-1969 ---------

e6_47_69 <- e6_47_69 %>% 
  filter(!is.na(`4/1/1940`)) %>% 
  mutate(County = ifelse(County == "California", "CALIFORNIA", County)) %>% 
  select(County, starts_with("7/")) %>% 
  pivot_longer(-County, names_to = "Year", values_to = "july1_population") %>% 
  mutate(Year = lubridate::mdy(Year), 
         Year = year(Year))

# All Other E6 files
 
processData <- function(myData, n = 10) {
  
  myData <- myData %>% 
    filter(!is.na(Year)) %>% 
    mutate(County = ifelse(County == "California", "CALIFORNIA", County))
  
  tList <- lapply(c("CALIFORNIA", countyLink$county), function(x) {
    countyIndex <- which(myData$County == x)
    myData[(countyIndex):(countyIndex+n), ] %>% 
      mutate(County = x)
  }) %>% 
    bind_rows() %>% 
    mutate(july1_population = as.numeric(july1_population))
    
  
  return(tList)
  
}

# Check county names --------

setdiff(unique(e6_70_80$County), countyLink$county)
setdiff(countyLink$county, unique(e6_70_80$County))

setdiff(unique(e6_80_90$County), countyLink$county)
setdiff(countyLink$county, unique(e6_80_90$County))

setdiff(unique(e6_80_90$County), countyLink$county)
setdiff(countyLink$county, unique(e6_80_90$County))

setdiff(unique(e6_90_00$County), countyLink$county)
setdiff(countyLink$county, unique(e6_90_00$County))

setdiff(unique(e6_00_10$County), countyLink$county)
setdiff(countyLink$county, unique(e6_00_10$County))

setdiff(unique(e6_10_21$County), countyLink$county)
setdiff(countyLink$county, unique(e6_10_21$County))

setdiff(unique(e6_20_22$County), countyLink$county)
setdiff(countyLink$county, unique(e6_20_22$County))

# Process data -----------
e6_70_80 <- processData(e6_70_80, 10) 

e6_80_90 <- processData(e6_80_90, 10) 

e6_90_00 <- processData(e6_90_00, 10) 

e6_00_10 <- processData(e6_00_10, 11) 

e6_10_21 <- processData(e6_10_21, 12) %>% 
  filter(!grepl("Census", Year)) %>% 
  mutate(Year = ifelse(Year == "Apr-Jun 2010", "2010", Year), 
         Year = as.numeric(Year)) 

e6_20_22 <- processData(e6_20_22, 3) %>% 
  filter(!grepl("Census", Year)) %>% 
  mutate(Year = ifelse(Year == "Apr-Jun 2020", "2020", Year), 
         Year = as.numeric(Year))

# Check for duplicates -------------

data.frame(Year = c(unique(e6_47_69$Year),
                    unique(e6_70_80$Year), 
                    unique(e6_80_90$Year), 
                    unique(e6_90_00$Year), 
                    unique(e6_00_10$Year), 
                    unique(e6_10_21$Year), 
                    unique(e6_20_22$Year)), stringsAsFactors = F) %>% 
  count(Year) %>% 
  filter(n > 1)

checkIfIdentical <- function(data1, data2, myYear) {
  
  data1 <- data1 %>% filter(Year == myYear) %>% arrange(County) 
  data2 <- data2 %>% filter(Year == myYear) %>% arrange(County)
  
  myCheck <- identical(
    data1, data2
  )
  
  if (myCheck) {
    print(TRUE)
  } else {
    full_join(rename(data1, pop1 = july1_population), 
              rename(data2, pop2 = july1_population)) %>% 
      mutate(diff = pop1 - pop2) %>% 
      View()
    
    
  }
}

checkIfIdentical(e6_70_80, e6_80_90, 1980)
checkIfIdentical(e6_80_90, e6_90_00, 1990) # FALSE
checkIfIdentical(e6_90_00, e6_00_10, 1999)
checkIfIdentical(e6_90_00, e6_00_10, 2000) # FALSE
checkIfIdentical(e6_00_10, e6_10_21, 2010) # FALSE
checkIfIdentical(e6_10_21, e6_20_22, 2020) # FALSE
checkIfIdentical(e6_10_21, e6_20_22, 2021) # FALSE

# e6 final -----------

e6_long <- e6_47_69 %>% 
  bind_rows(filter(e6_70_80, Year != 1980)) %>% 
  bind_rows(filter(e6_80_90, Year != 1990)) %>% 
  bind_rows(filter(e6_90_00, !Year %in% 1999:2000)) %>% 
  bind_rows(filter(e6_00_10, Year != 2010)) %>% 
  bind_rows(filter(e6_10_21, !Year %in% 2020:2021)) %>% 
  bind_rows(e6_20_22) %>% 
  rename(Population = july1_population)

table(e6_long$Year, useNA = "ifany")
colSums(is.na(e6_long))

e6_wide <- e6_long %>% 
  pivot_wider(names_from = Year, values_from = Population)

```

### E6_2000_2010 has estimates from 1999-2010. The excel spreadsheet says the 2010 Census benchmark is used, but the 2010 populations do not match the populations from the 2010 Census. It seems like DOF retroactively updated their estimates based on the 2010 Census. 

### Decision: Using July 1 E6 DOF estimates over Census estimates for all Decennial years to maintain consistency.

### Decision: Population estimates for most of the overlapping years in these E6 data files are different. Therefore, we are using the most recent estimate (meaning the estimate from the more recent decade)