---
title: "Exploration of 2020 Census Tracts - 2010 to 2020 mapping"
author: "Jaspreet Kang, Michael Samuel (OPP CDPH)"
date: "7/5/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

options(scipen = 999)

if (F) {
  server <- T
  if (server) setwd("/mnt/projects/FusionData/OTHER Analyses/2020 Census Tracts/")
  if (!server) setwd("G:/FusionData/OTHER Analyses/2020 Census Tracts/")
}

# Load packages -------------------------------------------------------------------------

library(readxl) # Read excel files
library(readr) # Read csv files
library(dplyr) # data wrangling
library(tidyr) # for pivoting
library(tibble)
library(tmap) # Creating maps
library(sf) # Simple features
library(tigris) 

# Load helpful info files ----------------------------------------------------------------

# county fips to county
countyLink <- read_xlsx("info/countyLink.xlsx") %>% 
  select(county = countyName, fips = FIPSCounty)

# Tract 2010 to 2020; downloaded from Census website
tract10_to_20 <- read_csv("info/tract10_to_tract20.csv") %>% 
  tibble::rowid_to_column("ID") %>% 
  mutate(fips20 = substr(GEOID_TRACT_20, 2, 4), 
         fips10 = substr(GEOID_TRACT_10, 2, 4), 
         across(starts_with("GEOID"), ~ paste0("0", .x))) %>% 
  left_join(rename(countyLink, county20 = county), by = c("fips20" = "fips")) %>% 
  left_join(rename(countyLink, county10 = county), by = c("fips10" = "fips")) 

# Load census TIGER/Line shapefiles using the tigris package
tract20_tigris <- tracts(state = 06, year = 2020) %>% 
  select(GEOID)
tract10_tigris <- tracts(state = 06, year = 2019) %>% 
  select(GEOID)


# Prepare data ----------------------------

# This data frame checks if areas (of land & water) are equivalent for each 2010, 2020 relationship
tract10_to_20_compareAreas <- tract10_to_20 %>% 
  select(ID, starts_with("county"), starts_with("GEOID"), starts_with("NAME"), starts_with("AREA")) %>% 
  mutate(is_equal_areaLand = if_all(c("AREALAND_TRACT_20", "AREALAND_TRACT_10"), `==`, AREALAND_PART), 
         is_equal_areaWater = if_all(c("AREAWATER_TRACT_20", "AREAWATER_TRACT_10"), `==`, AREAWATER_PART), 
         totalArea20 = AREALAND_TRACT_20 + AREAWATER_TRACT_20, 
         totalArea10 = AREALAND_TRACT_10 + AREAWATER_TRACT_10,
         compareArea = case_when(totalArea20 == totalArea10 ~ "==", 
                                 totalArea20 > totalArea10 ~ ">", 
                                 totalArea20 < totalArea10 ~ "<"))

```

# Links:

-   [Census' explanation of the 2020 Census Tract to 2010 Census Tract Relationship File](https://www2.census.gov/geo/pdfs/maps-data/data/rel2020/tract/explanation_tab20_tract20_tract10.pdf)

-   [Download relationship files from Census' website](https://www.census.gov/geographies/reference-files/time-series/geo/relationship-files.2020.html#list-tab-1709067297)

```{r}

length(unique(tract10_to_20$GEOID_TRACT_20))
length(unique(tract10_to_20$GEOID_TRACT_10)) 

```


```{r}
# 2020 Tracts with only one or multiple relationships -------------------------
tracts20_nRelationships <- tract10_to_20_compareAreas %>% 
  count(GEOID_TRACT_20)

tracts20_oneRelationship <- tracts20_nRelationships %>% 
  filter(n == 1) %>% 
  pull(GEOID_TRACT_20)

tracts20_multipleRelationships <- tracts20_nRelationships %>% 
  filter(n > 1) %>% 
  pull(GEOID_TRACT_20)
```

# 2020 Tracts with one relationship (only appears once in data)

## Identifying Possibility #1: 2020 Tracts where their boundaries did not change from 2010.

```{r}

identicalTracts <- tract10_to_20_compareAreas %>%
  filter(GEOID_TRACT_20 %in% tracts20_oneRelationship, 
         is_equal_areaLand, is_equal_areaWater) %>% 
  mutate(relationship = "Tract boundaries unchanged") %>% 
  select(ID, GEOID_TRACT_20, GEOID_TRACT_10, relationship)

identicalTracts_ID <- identicalTracts$ID

# Ensure these tracts appear only once
length(unique(identicalTracts_ID)) == nrow(identicalTracts) 


```


## Identifying Possibility #2: 2020 Tract wholly contained within a portion of a 2010 Tract

Steps:

- Filter on 2020 tracts with only one relationship in the data (only occurs once), then
- Exclude 2020 tracts where boundaries did nor change from 2010 (identified in previous R chunk)
- Check if total Area of 2020 Tracts < 2010 Tracts, which confirms a 2020 Tract is wholly contained within a portion of a 2010 Tract

```{r}

# Filter on 2020 tracts where boundaries changed from 2010, then Filter on 2020 tracts with only one relationship in the data
tracts20_whollyIn_portionTracts10 <- tract10_to_20_compareAreas %>% 
  filter(GEOID_TRACT_20 %in% tracts20_oneRelationship, 
         !ID %in% identicalTracts_ID 
         ) 

# Compare total areas -------------------------

# All total areas of 2020 tracts are < 2010 tracts, confirming these tracts are wholly contained within portions of 2010 tracts
table(tracts20_whollyIn_portionTracts10$compareArea)

tracts20_whollyIn_portionTracts10 <- tracts20_whollyIn_portionTracts10 %>% 
  mutate(relationship = "2020 Tract wholly contained within a portion of 2010 Tract") %>% 
  select(ID, GEOID_TRACT_20, GEOID_TRACT_10, relationship)
  
tracts20_whollyIn_portionTracts10_ID <- tracts20_whollyIn_portionTracts10$ID

```

```{r}
# Logic check
(length(identicalTracts_ID) + length(tracts20_whollyIn_portionTracts10_ID)) == length(tracts20_oneRelationship)


```

# 2020 Tracts with Multiple Relationships (appears multiple times in data)

## Possibility #3: Portion of 2020 Tract links to entire 2010 Tract

Steps:

- Filter on 2020 tracts where there are multiple relationships (2020 tract appears multiple times), then
- From this filtered data, pull 2010 tracts that only appear once, then
- From the filtered data from first step, filter on 2010 tracts that only appear once


```{r}

# Filter on 2020 tracts where there are multiple relationships
check <- tract10_to_20_compareAreas %>% 
  filter(GEOID_TRACT_20 %in% tracts20_multipleRelationships)

# Get count of 2010 tracts; pull 2010 tracts such that there is only one relationship
tracts10_oneRelationship <- check %>% 
  count(GEOID_TRACT_10) %>% 
  filter(n == 1) %>% 
  pull(GEOID_TRACT_10)

tractsPortion20_whollyIn_Tracts10 <- check %>% 
  filter(GEOID_TRACT_10 %in% tracts10_oneRelationship) %>% 
  mutate(relationship = "Portion of 2020 Tract links to entire 2010 Tract") %>% 
  select(ID, GEOID_TRACT_20, GEOID_TRACT_10, relationship)
  

tractsPortion20_whollyIn_Tracts10_ID <- tractsPortion20_whollyIn_Tracts10$ID

# Logic check ---------------------------------------------------------------------------------
# So far we've identified
# - 2020 tracts where boundaries did not change
# - 2020 tracts wholly contained within portion of 2010
# - 2010 tracts wholly contained within portion of 2020

length(identicalTracts_ID) + length(tracts20_whollyIn_portionTracts10_ID) + length(tractsPortion20_whollyIn_Tracts10_ID)

allIDs_identified <- c(identicalTracts_ID, tracts20_whollyIn_portionTracts10_ID, tractsPortion20_whollyIn_Tracts10_ID)

length(unique(allIDs_identified)) == length(allIDs_identified)

```

## Possibility #4: Portion of 2020 Tract in Portion of 2010 Tract

```{r}
# Filter on 2020 tracts where there are multiple relationships
check <- tract10_to_20_compareAreas %>% 
  filter(GEOID_TRACT_20 %in% tracts20_multipleRelationships)

# Get count of 2010 tracts; pull 2010 tracts such that there are multiple relationships
tracts10_multipleRelationships <- check %>% 
  count(GEOID_TRACT_10) %>% 
  filter(n > 1) %>% 
  pull(GEOID_TRACT_10)

tractsPortion20_in_TractsPortion10 <- check %>% 
  filter(GEOID_TRACT_10 %in% tracts10_multipleRelationships) %>% 
  mutate(relationship = "Portion of 2020 Tract in portion of 2010 Tract") %>% 
  select(ID, GEOID_TRACT_20, GEOID_TRACT_10, relationship)
  

tractsPortion20_in_TractsPortion10_ID <- tractsPortion20_in_TractsPortion10$ID


# Logic check ---------------------------------------------------------------------------------
# We should now have all relationships identified (language below needs more consistency)
# - 2020 tracts where boundaries did not change
# - 2020 tracts wholly contained within portion of 2010
# - Portion of 2020 tract that links to entire 2010 tract
# - Portion of 2020 tract in portion of 2010 tract

length(identicalTracts_ID) + length(tracts20_whollyIn_portionTracts10_ID) + length(tractsPortion20_whollyIn_Tracts10_ID) + length(tractsPortion20_in_TractsPortion10_ID)

allIDs_identified <- c(identicalTracts_ID, tracts20_whollyIn_portionTracts10_ID, tractsPortion20_whollyIn_Tracts10_ID, tractsPortion20_in_TractsPortion10_ID)

length(unique(allIDs_identified)) == length(allIDs_identified)


# All relationships are defined
finalShort <- bind_rows(identicalTracts, 
                        tracts20_whollyIn_portionTracts10,
                        tractsPortion20_whollyIn_Tracts10,
                        tractsPortion20_in_TractsPortion10) %>% 
  arrange(ID)

```

## Looking closer into possibility #3 

### EXPLORATORY CODE IN PROGRESS BELOW

```{r}

finalShort <- bind_rows(identicalTracts, 
                        tracts20_whollyIn_portionTracts10,
                        tractsPortion20_whollyIn_Tracts10,
                        tractsPortion20_in_TractsPortion10) %>% 
  arrange(ID)

# Should be no duplicate IDs
anyDuplicated(finalShort$ID)

geoids_p3 <- tractsPortion20_whollyIn_Tracts10 %>% 
  pull(GEOID_TRACT_20) %>% 
  unique()



check_p3 <- lapply(geoids_p3, function(x) {
  
  tDat <- finalShort %>% 
    filter(GEOID_TRACT_20 %in% x)
  
  if ( all(tDat$relationship == "Portion of 2020 Tract links to entire 2010 Tract") ) {
    tDat %>% 
      mutate(relationship = "2010 Tracts aggregate up to 2020 Tract")
  } else {
    tDAt
  }
  
  
}) 
  



```

# Final dataset with all relationships identified

```{r}


final <- tract10_to_20 %>% 
  left_join(finalShort, by = c("ID", "GEOID_TRACT_20", "GEOID_TRACT_10")) %>% 
  select(-c("ID", "fips20", "fips10"))

# No NAs should exist in relationship column
colSums(is.na(final))


table(finalShort$relationship, useNA = "ifany")

```


# Maps

```{r}

group_by_custom <- function(.data, myNewCol, ...) {
  .data %>% 
    group_by(...) %>% 
    mutate(!!as.symbol(myNewCol) := n()) %>% 
    ungroup() 
}

tDat <- final %>% 
  group_by_custom(myNewCol = "nTracts20", GEOID_TRACT_20) %>% 
  group_by_custom(myNewCol = "nTracts10", GEOID_TRACT_10) %>% 
  group_by_custom(myNewCol = "nRel20", GEOID_TRACT_20, relationship) %>% 
  group_by_custom(myNewCol = "nRel10", GEOID_TRACT_10, relationship) %>% 
  mutate(relationship2 = case_when(nTracts20 != nRel20 ~ ">1 relationship", 
                                   nTracts10 != nRel10 ~ ">1 relationship", 
                                   TRUE ~ relationship)) %>% 
  select(starts_with("GEOID"), starts_with("county"), starts_with("relationship"))


countyMap <- function(myCounty) {
  
  tDat <- tDat %>% 
    filter(county10 == myCounty, county20 == myCounty)
  
  tDat10 <- tDat %>% 
    select(GEOID = GEOID_TRACT_10, relationship2) %>% 
    distinct(GEOID, .keep_all = TRUE)
  mDat10 <- tract10_tigris %>% 
    right_join(tDat10)
  
  tDat20 <- tDat %>% 
    select(GEOID = GEOID_TRACT_20, relationship2) %>% 
    distinct(GEOID, .keep_all = TRUE)
  mDat20 <- tract20_tigris %>% 
    right_join(tDat20)
  
  map10 <- tm_shape(mDat10) +
    tm_polygons("relationship2", border.col = "blue", lwd = 1) +
    tm_layout(title = "2010 Tracts")
  
  map20 <- tm_shape(mDat20) +
    tm_polygons("relationship2", border.col = "blue", lwd = 1) +
    tm_layout(title = "2020 Tracts")
  
  tmap_mode("view")
  tmap_arrange(map10, map20, ncol = 1, sync = TRUE)
  

}

countyMap("Yolo")



```

